---
title: "Exercise 10"
format: html
---

## Step 1

Load needed LIBRARIES

```{r}
library(readr)
library(dplyr)
library(tidyr)
library(skimr)
library(ggplot2)
library(cowplot)
library(infer)
```

DATA
```{r}
f <- "https://raw.githubusercontent.com/difiore/ada-datasets/refs/heads/main/AVONETdataset1.csv" #file path
d <- read_csv(f, col_names = TRUE) #read dataset as tibble
```

Clean data
```{r}
d <- d |>
  select(Species1, Family1, Order1, Beak.Length_Culmen, Beak.Width, Beak.Depth, Tarsus.Length, Wing.Length, Tail.Length, Mass, Habitat, Migration, Trophic.Level, Trophic.Niche, Min.Latitude, Max.Latitude, Centroid.Latitude, Primary.Lifestyle, Range.Size)
```

```{r}
#create log mass variable
d <- d |> mutate(logMass = log(Mass))
```

```{r}
# Boxplot: logMass by Trophic.Level
ggplot(d, aes(x = Trophic.Level, y = logMass)) +
  geom_boxplot() +
  labs(title = "Log(Mass) by Trophic Level",
       x = "Trophic Level",
       y = "Log(Mass)") +
  theme_minimal()

# Convert Migration to a factor with meaningful labels
d <- d |> 
  mutate(Migration = factor(Migration,
                            levels = c(1, 2, 3),
                            labels = c("Resident", "Partial Migrant", "Migrant")))

# Then replot with the corrected factor levels
ggplot(d, aes(x = Migration, y = logMass)) +
  geom_boxplot() +
  labs(title = "Log(Mass) by Migration Behavior",
       x = "Migration Type",
       y = "Log(Mass)") +
  theme_minimal()
```
## Step 2

```{r}
# Linear model: logMass ~ Trophic.Level
mod_trophic <- lm(logMass ~ Trophic.Level, data = d)
summary(mod_trophic)

# Linear model: logMass ~ Migration
mod_migration <- lm(logMass ~ Migration, data = d)
summary(mod_migration)

```
The f-stat for  both models is large and significant with the p values also being small -> we reject null hypothesis

```{r}
# Change reference level to "Partial Migrant"
d <- d |> mutate(Migration = relevel(Migration, ref = "Partial Migrant"))

# Run the model again
mod_migration_relevel <- lm(logMass ~ Migration, data = d)
summary(mod_migration_relevel)
```
Partial Migrants are the heaviest group on average (since they are the refernce).

Residents are significantly lighter than Partial Migrants by ~0.76 log units.

Migrants are also significantly lighter than Partial Migrants by ~0.38 log units.

## Step 3

```{r}
anova_migration <- aov(logMass ~ Migration, data = d)

tukey_migration <- TukeyHSD(anova_migration)
tukey_migration
```
## Step 4

```{r}
# Step 1: Start with your dataset (remove missing values to match lm())
d_clean <- d |> filter(!is.na(logMass), !is.na(Trophic.Level))

# Step 2: Specify the model and simulate null distribution
null_distribution <- d_clean |>
  specify(logMass ~ Trophic.Level) |>
  hypothesize(null = "independence") |>
  generate(reps = 1000, type = "permute") |>
  calculate(stat = "F")

# Step 3: Calculate the observed F-statistic from your original model
observed_f <- lm(logMass ~ Trophic.Level, data = d_clean) |> 
  summary() |> 
  (\(x) x$fstatistic[1])() |> 
  as.numeric()

# Step 4: Visualize null distribution and add observed F
ggplot(null_distribution, aes(x = stat)) +
  geom_histogram(bins = 30, fill = "lightsteelblue", color = "white") +
  geom_vline(xintercept = observed_f, color = "hotpink2", linetype = "dashed") +
  labs(title = "Permutation Test: F-statistic for Trophic.Level",
       x = "F-statistic (null distribution)",
       y = "Count") +
  theme_minimal()

# Step 5: Calculate empirical p-value
p_value <- null_distribution |>
  summarize(p_val = mean(stat >= observed_f))

p_value

```

# Challenge 2

## Step 1
```{r}
# Calculate relative beak length (residuals from log-log regression)
relBeak <- lm(log(Beak.Length_Culmen) ~ logMass, data = d)
d <- d |> mutate(relBeak = resid(relBeak))
```

```{r}
# Calculate relative tarsus length (residuals from log-log regression)
relTarsus <- lm(log(Tarsus.Length) ~ logMass, data = d)
d <- d |> mutate(relTarsus = resid(relTarsus))
```

```{r}
# Boxplot 1: relTarsus by Primary.Lifestyle
p1 <- ggplot(data = d |> drop_na(Primary.Lifestyle),
             aes(x = Primary.Lifestyle, y = relTarsus)) +
  geom_boxplot() +
  labs(title = "Relative Tarsus Length by Primary Lifestyle",
       x = "Primary Lifestyle", y = "Relative Tarsus Length") +
  theme_minimal()

p1

# Boxplot 2: relBeak by Trophic.Niche
p2 <- ggplot(data = d |> drop_na(Trophic.Niche),
             aes(x = Trophic.Niche, y = relBeak)) +
  geom_boxplot() +
  labs(title = "Relative Beak Length by Trophic Niche",
       x = "Trophic Niche", y = "Relative Beak Length") +
  theme_minimal() 

# Arrange the two plots side-by-side
plot_grid(p1, p2, nrow = 1)
```
## Step 3

```{R}
# Histogram of Range.Size
ggplot(d, aes(x = Range.Size)) +
  geom_histogram(bins = 50, fill = "lightsteelblue", color = "white") +
  scale_x_continuous(trans = "log10") +
  labs(title = "Distribution of Range Size (log10 scale)", x = "Range Size", y = "Count") +
  theme_minimal()

```

```{r}
# Drop NAs and transform Range.Size
d_range <- d |> 
  filter(!is.na(Migration), !is.na(Range.Size)) |>
  mutate(logRange = log(Range.Size))  # log10 is common for range area

# Make sure Migration is a factor (should already be from previous steps)
d_range$Migration <- factor(d_range$Migration)

# Run ANOVA
mod_range <- aov(logRange ~ Migration, data = d_range)
summary(mod_range)

```

## Step 4
```{r}
pass <- d |> filter(Order1 == "Passeriformes")
```

```{r}
# One-factor models (separate)
m1 <- aov(relBeak ~ Primary.Lifestyle, data = pass)
m2 <- aov(relBeak ~ Trophic.Level, data = pass)

# Summary of each
summary(m1)
summary(m2)
```

``` {r}
m1 <- aov(relBeak ~Primary.Lifestyle + Trophic.Level, data = pass)
m2 <- aov(relTarsus ~Primary.Lifestyle + Trophic.Level, data = pass)
```

```{r}
# Boxplot: relBeak ~ Primary.Lifestyle
p1 <- ggplot(pass, aes(x = Primary.Lifestyle, y = relBeak)) +
  geom_boxplot() +
  labs(title = "Relative Beak Length by Primary Lifestyle") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# Boxplot: relBeak ~ Trophic.Level
p2 <- ggplot(pass, aes(x = Trophic.Level, y = relBeak)) +
  geom_boxplot() +
  labs(title = "Relative Beak Length by Trophic Level") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# Boxplot: relBeak by combination of both predictors
p3 <- ggplot(pass, aes(x = interaction(Primary.Lifestyle, Trophic.Level), y = relBeak)) +
  geom_boxplot() +
  labs(title = "Relative Beak Length by Lifestyle & Trophic Level",
       x = "Lifestyle.Trophic") +
  theme(axis.text.x = element_text(angle = 60, hjust = 1))

# Arrange plots side-by-side
plot_grid(p1, p2, p3, nrow = 1)
```


## Step 5 

2-way model adding the possibilty of an interaction effect; to do this, we would modify our model formula a bit using the colo operator (:) to specify a particular interaction term or the asterisk operator (*) to specify all single variables + interaction terms ...

```{r}
m5 <- aov(relBeak ~ Primary.Lifestyle + Trophic.Level, data = pass)
summary(m5)
```

```{r}
m8 <- aov(relBeak ~Primary.Lifestyle + Trophic.Level + Primary.Lifestyle:Trophic.Level, data = pass)
summary(m8)
```

## Step 7

```{r}
# Make sure the factors are set correctly
pass$Primary.Lifestyle <- factor(pass$Primary.Lifestyle)
pass$Trophic.Level <- factor(pass$Trophic.Level)

# Create the interaction plot
interaction.plot(x.factor = pass$Trophic.Level,
                 trace.factor = pass$Primary.Lifestyle,
                 response = pass$relBeak,
                 fun = mean,
                 type = "b",
                 legend = TRUE,
                 xlab = "Trophic Level",
                 ylab = "Mean Relative Beak Length",
                 trace.label = "Primary Lifestyle",
                 col = 1:6,
                 pch = 1:6,
                 lty = 1,
                 main = "Interaction Plot: Primary Lifestyle × Trophic Level")
```
## Step 8

```{r}
# From model with relBeak ~ Trophic.Level
mod_check <- lm(relBeak ~ Trophic.Level, data = pass)
pass$resid_check <- resid(mod_check)
pass$fitted_check <- fitted(mod_check)

ggplot(pass, aes(x = resid_check)) +
  geom_histogram(bins = 30, fill = "lightsteelblue", color = "white") +
  facet_wrap(~Trophic.Level, scales = "free") +
  labs(title = "Distribution of Residuals by Trophic Level",
       x = "Residuals", y = "Count")

```
The residuals appear approximately normally distributed within each trophic level, which supports the assumption of normality.


Class Notes
```{r}
install.packages("car")
install.packages("jtools")
library(car)
library(jtools)
```

```{r}
f <- "https://raw.githubusercontent.com/difiore/ada-datasets/refs/heads/main/zombies.csv"

z <- read_csv(f, col_names = TRUE)

m <- lm(height ~ weight + age, data = z)
summary(m)

```

```{r}
plot(m$model$weight, residuals(m)) #relationship of each predictor and residuals of model
plot(m$model$age, residuals(m))
plot(fitted(m), residuals(m))
summary(aov(m))
```

```{r}
(f <- (summary(m)$r.squared*(nrow(z)-2-1))/((1-summary(m)$r.squared)*2)) #calculate f statistic
```
```{r}
temp <- lm(height ~ weight + age + gender, data = z)
summary(m)
```

```{r}
boxplot(residuals(m) ~ m$model$gender)
```
Variance inflation factor is how easily each predictor variable is predicted by other predicted variables
to what extent is gender predicted by age and weight
to what extent is age predicted by gender and weight
to what extent is weight predicted by gender and age

```{r}
vif(m)
```







